---
subtitle: "TMA4268 Statistical Learning V2019"
title: "Compulsory exercise 1: Group 2"
author: "Anders Thallaug Fagerli, Rebecca Sandstø"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document
  #pdf_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(results = "hold")
knitr::opts_chunk$set(error=FALSE)
knitr::opts_chunk$set(warning=FALSE)
```

# Problem 1: Multiple linear regression 

```{r,echo=TRUE,eval=TRUE}
library(GLMsData)
data("lungcap")
lungcap$Htcm=lungcap$Ht*2.54
modelA = lm(log(FEV) ~ Age + Htcm + Gender + Smoke, data=lungcap)
summary(modelA)
```

**Q1:** Write down the equation for the fitted modelA.
$**Y** = **X**B

**Q2:** 
    
* `Estimate` - in particular interpretation of `Intercept`
* `Std.Error`
* `Residual standard error`
* `F-statistic`

**Q3:** 

**Q4:**

```{r,eval=TRUE}
library(ggplot2)
# residuls vs fitted
ggplot(modelA, aes(.fitted, .stdresid)) + geom_point(pch = 21) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(se = FALSE, col = "red", size = 0.5, method = "loess") +
  labs(x = "Fitted values", y = "Standardized residuals",
       title = "Fitted values vs. Standardized residuals",
       subtitle = deparse(modelA$call))

# qq-plot of residuals
ggplot(modelA, aes(sample = .stdresid)) +
  stat_qq(pch = 19) + 
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  labs(x = "Theoretical quantiles", y = "Standardized residuals", 
       title = "Normal Q-Q", subtitle = deparse(modelA$call))

# normality test
library(nortest) 
ad.test(rstudent(modelA))
```

**Q5:** 

```{r,eval=TRUE}
modelB = lm(FEV ~ Age + Htcm + Gender + Smoke, data=lungcap)
summary(modelB)
```

**Q6:** 

```{r,eval=TRUE}
# here you write your code if you have any
```

**Q7:** 

```{r,eval=TRUE}
# here you write your code
```


**Q8:**

```{r,eval=TRUE}
new = data.frame(Age=16, Htcm=170, Gender="M", Smoke=0)
```

# Problem 2: Classification 

```{r}
library(class)# for function knn
library(caret)# for confusion matrices

raw = read.csv("https://www.math.ntnu.no/emner/TMA4268/2019v/data/tennis.csv")
M = na.omit(data.frame(y=as.factor(raw$Result),
                       x1=raw$ACE.1-raw$UFE.1-raw$DBF.1, 
                       x2=raw$ACE.2-raw$UFE.2-raw$DBF.2))
set.seed(4268) # for reproducibility
tr = sample.int(nrow(M),nrow(M)/2)
trte=rep(1,nrow(M))
trte[tr]=0
Mdf=data.frame(M,"istest"=as.factor(trte))
```

**Q9:** 


**Q10:** 

```{r,eval=TRUE}
# here you write your code
ks = 1:30
yhat = sapply(ks, function(k) {
  class::knn(train = M[tr,-1], cl = M[tr,1],test = M[-1], k = k)
})
train.equal = I(yhat == M[,1])
train.e = sum(train.equal[,1] == FALSE)/length(train.equal[,1])
yhat = sapply(ks, function(k) {
  class::knn(train = M[-tr,-1], cl = M[-tr,1],test = M[-1], k = k)
})
test.equal = I(yhat == M[,1])
test.e = sum(test.equal[,1] == FALSE)/length(test.equal[,1])

```


```{r, eval=FALSE}
set.seed(0)
ks = 1:30 # Choose K from 1 to 30.
idx = createFolds(M[tr,1], k=5) # Divide the training data into 5 folds.
# "Sapply" is a more efficient for-loop. 
# We loop over each fold and each value in "ks"
# and compute error rates for each combination.
# All the error rates are stored in the matrix "cv", 
# where folds are rows and values of $K$ are columns.
cv = sapply(ks, function(k){ 
  sapply(seq_along(idx), function(j) {
    yhat = class::knn(train=M[tr[ -idx[[j]] ], -1],
               cl=M[tr[ -idx[[j]] ], 1],
               test=M[tr[ idx[[j]] ], -1], k = k)
    mean(M[tr[ idx[[j]] ], 1] != yhat)
  })
})
```


**Q11:** 

```{r, eval=FALSE}
cv.e = numeric(30) # fill in
for (i in 1:30) {
  cv.e[i] <- mean(cv[,i])
}
cv.se = sd(cv)/sqrt(5) #fill in
k.min = col(cv)[cv==min(cv)] # fill in
```

**Q12:** 

```{r,eval=FALSE}
library(colorspace)
co = rainbow_hcl(3)
par(mar=c(4,4,1,1)+.1, mgp = c(3, 1, 0))
plot(ks, cv.e, type="o", pch = 16, ylim = c(0, 0.7), col = co[2],
     xlab = "Number of neighbors", ylab="Misclassification error")
arrows(ks, cv.e-cv.se, ks, cv.e+cv.se, angle=90, length=.03, code=3, col=co[2])
lines(ks, train.e, type="o", pch = 16, ylim = c(0.5, 0.7), col = co[3])
lines(ks, test.e, type="o", pch = 16, ylim = c(0.5, 0.7), col = co[1])
legend("topright", legend = c("Test", "5-fold CV", "Training"), lty = 1, col=co)
```

**Q13:** 

```{r,eval=FALSE}
k = tail(which(cv.e < cv.e[k.min] + cv.se[k.min]), 1)
size = 100
xnew = apply(M[tr,-1], 2, function(X) seq(min(X), max(X), length.out=size))
grid = expand.grid(xnew[,1], xnew[,2])
grid.yhat = knn(M[tr,-1], M[tr,1], k=k, test=grid)
np = 300
par(mar=rep(2,4), mgp = c(1, 1, 0))
contour(xnew[,1], xnew[,2], z = matrix(grid.yhat, size), levels=.5, 
        xlab=expression("x"[1]), ylab=expression("x"[2]), axes=FALSE,
        main = paste0(k,"-nearest neighbors"), cex=1.2, labels="")
points(grid, pch=".", cex=1, col=grid.yhat)
points(M[1:np,-1], col=factor(M[1:np,1]), pch = 1, lwd = 1.5)
legend("topleft", c("Player 1 wins", "Player 2 wins"), 
       col=c("red", "black"), pch=1)
box()
```


**Q14:** 

```{r,eval=FALSE}
K=# your choice from Q13
  
# knn with prob=TRUE outputs the probability of the winning class
# therefore we have to do an extra step to get the probability of player 1 winning
KNNclass=class::knn(train=M[tr,-1], cl=M[tr,1], test=M[-tr,-1], k = K,prob=TRUE)
KNNprobwinning=attributes(KNNclass)$prob
KNNprob= ifelse(KNNclass == "0", 1-KNNprobwinning, KNNprobwinning)
# now KNNprob has probability that player 1 wins, for all matches in the test set

library(pROC)
# now you use predictor=KNNprob and response=M[-tr,1] 
# in your call to the function roc in the pROC library
```

**Q15:**

```{r,eval=TRUE}
# here you write your code
```


# Problem 3: Bias-variance trade-off 

Here you see how to write formulas with latex (needed below)
$$
\hat{\boldsymbol \beta}=({\bf X}^T{\bf X})^{-1}{\bf X}^T{\bf Y}
$$

**Q16:** 

**Q17:** 

**Q18:** 
$$\text{E}[(Y_0-\hat{f}({\bf x}_0))^2]=[\text{E}(\hat{f}({\bf x}_0)-f({\bf x}_0)]^2+\text{Var}(\hat{f}({\bf x}_0) ) + \text{Var}(\varepsilon)$$

Ridge estimator:
$$
\widetilde{\boldsymbol \beta}=({\bf X}^T{\bf X}+\lambda {\bf I})^{-1}{\bf X}^T{\bf Y}
$$

**Q19:** 

**Q20:** 

**Q21:** 
$$\text{E}[(Y_0-\widetilde{f}({\bf x}_0))^2]=[\text{E}(\widetilde{f}({\bf x}_0)-f({\bf x}_0)]^2+\text{Var}(\widetilde{f}({\bf x}_0) ) + \text{Var}(\varepsilon)$$


```{r}
values=dget("https://www.math.ntnu.no/emner/TMA4268/2019v/data/BVtradeoffvalues.dd")
X=values$X
dim(X)
x0=values$x0
dim(x0)
beta=values$beta
dim(beta)
sigma=values$sigma
sigma
```

Hint: we perform matrix multiplication using `%*%`, transpose of a matrix `A` with `t(A)` and inverse with `solve(A)`. 

**Q22:** 

```{r,eval=FALSE}
sqbias=function(lambda,X,x0,beta)
{
  p=dim(X)[2]
  value= #HERE YOU FILL IN
  return(value)
}
thislambda=seq(0,2,length=500)
sqbiaslambda=rep(NA,length(thislambda))
for (i in 1:length(thislambda)) sqbiaslambda[i]=sqbias(thislambda[i],X,x0,beta)
plot(thislambda,sqbiaslambda,col=2,type="l")
```

**Q23:** 

```{r,eval=FALSE}
variance=function(lambda,X,x0,sigma)
{
  p=dim(X)[2]
  inv=solve(t(X)%*%X+lambda*diag(p))
  value=#HERE YOU FILL IN
  return(value)
}
thislambda=seq(0,2,length=500)
variancelambda=rep(NA,length(thislambda))
for (i in 1:length(thislambda)) variancelambda[i]=variance(thislambda[i],X,x0,sigma)
plot(thislambda,variancelambda,col=4,type="l")
```


**Q24:** 

```{r,eval=FALSE}
tot=sqbiaslambda+variancelambda+sigma^2
which.min(tot)
thislambda[which.min(tot)]
plot(thislambda,tot,col=1,type="l",ylim=c(0,max(tot)))
lines(thislambda, sqbiaslambda,col=2)
lines(thislambda, variancelambda,col=4)
lines(thislambda,rep(sigma^2,500),col="orange")
abline(v=thislambda[which.min(tot)],col=3)
```
